DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life

### 2023-02-21 manuscript/manuscript.Rmd
format response as md file

**_Dating a tree with branch lengths.--_**
Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain markers for input taxa.
<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
By default, mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa]. Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing]. The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn]. 
The BOLD workflow is currently only available on DateLife's R package and not on the web application.

@@

Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>). The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1#readme>, <https://github.com/LunaSare/datelife_examples#readme>, and <https://github.com/LunaSare/datelife_benchmark#readme>.

>>>>

**_Dating a tree with branch lengths.--_**
Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
DateLife implements a workflow to streamline this process by leveraging on open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
By default, BOLD genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa]. Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing]. The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site. Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn]. 
Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.

@@

Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>). The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
